---
title: "DAT301 Porfolio 2"
author: "Chijiuka Fortune Akuma"
Title: "Data Analysis and Visulaisation"
Student_Number: "2423891"
date:  "2025-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading of Necessary Libraries

```{r}
      library(tseries)
      library(zoo)
      library(forecast)
      library(dplyr)
      library(xts)
      library(tidyverse)
      library(ggplot2)
      library(readr)
      library(readxl)
      library(lubridate)
      library(tibble)
      library(lmtest)
      library(vars)
      library(Metrics)
      library(MTS)
      library(prophet)
      library(keras)
      library(tensorflow)
      library(corrplot)
      library(randomForest)
      library(rpart)
      library(caret)
      library(e1071)
      library(pROC)
      library(reticulate)
      library(doParallel)
      library(mlr)
      library(stats)
      library(ggpubr)
      library(cowplot)
      library(car)
      library(broom)
      library(aod)
      library(scales)
      library(RColorBrewer)
      library(reshape2)
      library(effsize)
      library(plotly)
```

# Load the Datasets

## Data-set 1

```{r}
    #SETTING THE WORK DIRECTORIES 
      setwd("C:/Users/-/OneDrive - University of Bolton/Eugene")
      
      ###LOADING OF THE DATASETS ----------------
      ## INFLATION RATES DATASET------------
      inflation_rate <- read.csv("NigeriaInflationRates - Copy.csv")
```

## Inspect the Data-set

```{r}
# Inspect dataset
      head(inflation_rate)
      view(inflation_rate)
      summary(inflation_rate)
```

## Convert the column to Date format

```{r}
 # Convert the column to Date format
      inflation_rate$Date <- as.Date(inflation_rate$Date, format = "%d/%m/%Y")
      
```

## Drop columns that are not important

```{r}
# Drop columns that are not important
      inflation_rate <- dplyr::select(inflation_rate, -Production, -Crude.Oil.Export)
```

## Convert the variables to appropriate datatypes

```{r}
#Using Functions to convert the variables to appropriate datatypes
      clean_inflation_data <- function(df) {
        df$Date <- as.Date(df$Date, format = "%d/%m/%Y")
        
        numeric_cols <- c(
          "Inflation_Rate", "Crude.Oil.Price", "CPI_Food", "CPI_Energy",
          "CPI_Health", "CPI_Transport", "CPI_Communication", "CPI_Education"
        )
        
        df[numeric_cols] <- lapply(df[numeric_cols], as.numeric)
        
        return(df)
      }
      
      inflation_rate <- clean_inflation_data(inflation_rate)
```

## Convert the Date column to index Column

```{r}
### Converting the Date column to index Column 
      inflation_rate <- column_to_rownames(inflation_rate, var = "Date")
```

## Detect NAs, NULL and Missing Values

```{r}
## Checking for NAs, NULL and Missing Values
      check_missing_values <- function(df) {
        cat("=== Missing Values Summary ===\n")
        
        # Check for NA values per column
        na_counts <- sapply(df, function(x) sum(is.na(x)))
        cat("\nNA values per column:\n")
        print(na_counts[na_counts > 0])
        
        # Check for NULL values per column (not common inside data frames)
        null_counts <- sapply(df, function(x) sum(sapply(x, is.null)))
        cat("\nNULL values per column:\n")
        print(null_counts[null_counts > 0])
        
        # Total NAs and NULLs
        total_na <- sum(is.na(df))
        total_null <- sum(unlist(sapply(df, function(x) sum(sapply(x, is.null)))))
        
        cat("\nTotal NA values in dataset:", total_na, "\n")
        cat("Total NULL values in dataset:", total_null, "\n")
      }
      ## Apply the function to the Dataset
      check_missing_values(inflation_rate)
```

## Handle detected NAs,NULL, or Missing values

```{r}
##Using functions to Handle detected NAs,NULL, or MIssing values
      impute_missing_with_approx <- function(df) {
        cols_with_na <- names(df)[colSums(is.na(df)) > 0]
        
        for (col in cols_with_na) {
          if (is.numeric(df[[col]])) {
            df[[col]] <- na.approx(df[[col]], na.rm = FALSE)
            cat(paste("Interpolated missing values in:", col, "\n"))
          } else {
            cat(paste("Skipped non-numeric column:", col, "\n"))
          }
        }
        
        return(df)
      }
      ##Apply the function on the dataset
      inflation_rate <- impute_missing_with_approx(inflation_rate)
```

# Load Data-set 2

```{r}
##LOADING OF SECOND DATASET (EXCHANGE RATE) 
    xchange_rate <- read.csv("USD_NGN Historical Data.csv")
```

## Inspect the Dataset

```{r}
head(xchange_rate)
      view(xchange_rate)
```

## Drop columns that are not important

```{r}
# Drop columns that are not important
      xchange_rate <- dplyr::select(xchange_rate, -Open, -High, -Low, -Vol., -Change..)
```

## Convert the Columns into appropriate data types

```{r}
##Convert the columns into approprate data types
      xchange_rate$Date <- as.Date(xchange_rate$Date, format="%m/%d/%Y")
      xchange_rate$Price <- as.numeric(xchange_rate$Price)
      str(xchange_rate)
      view(xchange_rate)
      # Check if there are any non-date values in the Date column
      invalid_dates <- xchange_rate[!grepl("^\\d{4}-\\d{2}-\\d{2}$", xchange_rate$Date), ]
      
      # Print invalid dates
      print(invalid_dates)
```

## Check for NAs, NULL and Missing Values

```{r}
## Checking for NAs, NULL and Missing Values
      xchange_missing_values <- function(df) {
        cat("=== Missing Values Summary ===\n")
        
        # Check for NA values per column
        na_counts <- sapply(df, function(x) sum(is.na(x)))
        cat("\nNA values per column:\n")
        print(na_counts[na_counts > 0])
        
        # Check for NULL values per column (not common inside data frames)
        null_counts <- sapply(df, function(x) sum(sapply(x, is.null)))
        cat("\nNULL values per column:\n")
        print(null_counts[null_counts > 0])
        
        # Total NAs and NULLs
        total_na <- sum(is.na(df))
        total_null <- sum(unlist(sapply(df, function(x) sum(sapply(x, is.null)))))
        
        cat("\nTotal NA values in dataset:", total_na, "\n")
        cat("Total NULL values in dataset:", total_null, "\n")
      }
      ## Apply the function to the Dataset
      xchange_missing_values(xchange_rate)
```

## Handle detected NAs,NULL, or Missing values

```{r}
##Using functions to Handle detected NAs,NULL, or MIssing values
      impute_missing_values <- function(df) {
        cols_with_na <- names(df)[colSums(is.na(df)) > 0]
        
        for (col in cols_with_na) {
          if (is.numeric(df[[col]])) {
            df[[col]] <- na.approx(df[[col]], na.rm = FALSE)
            cat(paste("Interpolated missing values in:", col, "\n"))
          } else {
            cat(paste("Skipped non-numeric column:", col, "\n"))
          }
        }
        
        return(df)
      }
      ##Apply the function on the dataset
      xchange_rate <- impute_missing_values(xchange_rate)
```

## Convert the Daily Exchange Rate to Monthly Data

```{r}
 ## Convert the Daily Exchange Rate to Monthly Data
      
      xchange_rate <- xchange_rate %>%
        mutate(Month = floor_date(Date, "month")) %>%  # Convert to start of the month
        group_by(Month) %>%  # Group by the new Month column
        summarize(Monthly_Average = mean(Price, na.rm = TRUE))  # Calculate average price for each month
      view(xchange_rate)
```

## Rename the columns in xchange_rate dataset

```{r}
# Rename the columns in xchange_rate dataset
      xchange_rate <- xchange_rate %>%
        rename(Date = Month, Price = Monthly_Average)  # Rename Month -> Date and Monthly_Average -> Price
```

## Filter data from 2008 to 2024

```{r}
# Filter data from 2008 to 2024
      xchange_rate_filtered <- xchange_rate %>%
        filter(year(Date) >= 2008 & year(Date) <= 2024)
```

## View the dataset

```{r}
# View the dataset
      print(xchange_rate_filtered)
      view(xchange_rate_filtered)
```

# COMBINE THE TWO DATASETS

## Verify the length of the two data sets

```{r}
# Verify the length of the two data sets
      nrow(xchange_rate_filtered)
      nrow(inflation_rate)
```

## Drop the last row from the Inflate Rate dataset to match the Exchange rate dataset

```{r}
 # Drop the last row from the Inflate Rate dataset to match the Exchange rate dataset
      inflation_rate <- inflation_rate %>%
        slice(-n())  # Exclude the last row
```

## Merge the data

```{r}
#Add the monthly  exchange rate to the inflation rate dataset
      inflation_rate$`Exchange Rate` <- xchange_rate_filtered$Price
```

## Summary of the dataset

```{r}
##SUMMARY OF THE DATSET
      inf_rate <- inflation_rate
```

## Description of the dataset

```{r}
 
      # Enhanced stats
      psych::describe(inf_rate)
      
      view(inf_rate)
```

# EXPLORATORY ANALYSIS

## Summary Statistics

```{r}
psych::describe(inf_rate[, c("CPI_Food", "CPI_Energy", "CPI_Health", 
                                   "CPI_Transport", "CPI_Communication", 
                                   "CPI_Education")])
```

# Impact of Inflation Rate on CPI Categories

```{r}
cpi_vars <- c("CPI_Food", "CPI_Energy", "CPI_Health", 
                    "CPI_Transport", "CPI_Communication", "CPI_Education")
      
      # Initialize a list to store results
      reg_results <- lapply(cpi_vars, function(var) {
        formula <- as.formula(paste(var, "~ Inflation_Rate"))
        model <- lm(formula, data = inf_rate)
        summary(model)$coefficients[2,1]  # get slope (beta)
      })
      
      impact_df <- data.frame(Feature = cpi_vars, 
                              Slope = unlist(reg_results))
```

## Visualise the Impact

```{r}
ggplot(impact_df, aes(x = reorder(Feature, Slope), y = Slope, fill = Slope)) +
        geom_bar(stat = "identity") +
        coord_flip() +
        labs(title = "Impact of Inflation Rate on CPI Categories",
             x = "CPI Category", y = "Estimated Impact (Slope Coefficient)") +
        theme_minimal(base_size = 14) +
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0)
```

## Correlation Heatmap: Inflation Rate vs. CPI Categories

```{r}
# All variables: Inflation + CPIs
      cpis <- c("Inflation_Rate", 
                    "CPI_Food", "CPI_Energy", "CPI_Health", 
                    "CPI_Transport", "CPI_Communication", "CPI_Education")
      
      cpi_data <- inf_rate[, cpis]
      
      # Compute correlation matrix
      cor_matrix <- cor(cpi_data, use = "complete.obs")
```

## Visualise the Heatmap

```{r}
# Create a clean, high-contrast heatmap
      corrplot(cor_matrix,
               method = "color",            # use color tiles
               type = "upper",              # show upper triangle
               col = colorRampPalette(c("darkred", "white", "darkblue"))(100),  # fine-grained diverging color
               addCoef.col = "white",       # add black correlation coefficients
               number.cex = 0.8,            # font size for numbers
               tl.col = "black",            # axis text color
               tl.srt = 45,                 # axis text rotation
               diag = FALSE,                # hide diagonal
               addgrid.col = "grey",        # grid line color
               mar = c(0, 0, 2, 0),         # margins
               title = "Correlation Heatmap: Inflation Rate vs. CPI Categories")
```

# Impact of Crude Oil Price and Exchange Rate on Inflation Rate

```{r}
# Fit a multiple linear regression model
      model_inflation_impact <- lm(Inflation_Rate ~ Crude.Oil.Price + `Exchange Rate`, data = inf_rate)
```

## Summary of the model to see the coefficients and statistical significance

```{r}
      # Summary of the model to see the coefficients and statistical significance
      summary(model_inflation_impact)
```

## Extract coefficients for Crude Oil Price and Exchange Rate

```{r}
      # Extract coefficients for Crude Oil Price and Exchange Rate
      coefficients_df <- data.frame(
        Feature = c("Crude Oil Price", "Exchange Rate"),
        Coefficient = coef(model_inflation_impact)[2:3]  # Get coefficients for Crude Oil Price and Exchange Rate
      )
```

## Visualise the Impact

```{r}
# Create a clean and well-aligned bar plot with a bold vertical line at 0 on the y-axis
      ggplot(coefficients_df, aes(x = reorder(Feature, Coefficient), y = Coefficient, fill = Coefficient)) +
        geom_bar(stat = "identity", color = "black", width = 0.75, show.legend = FALSE) +  # Clean bars
        coord_flip() +  # Flip coordinates to make labels readable
        labs(
          title = "Impact of Crude Oil Price and Exchange Rate on Inflation Rate",
          x = NULL,  # Remove x-axis label for cleaner look
          y = "Coefficient (Impact on Inflation Rate)"
        ) +
        theme_minimal(base_size = 16) +  # Use a larger base font size for clarity
        theme(
          plot.title = element_text(hjust = 0.5, face = "bold", size = 18),  # Center title and bold
          axis.text.y = element_text(size = 14),  # Adjust y-axis label font size
          axis.text.x = element_text(size = 12),  # Adjust x-axis label font size
          axis.title.y = element_text(size = 14),  # Make y-axis title slightly larger
          axis.title.x = element_blank(),  # No x-axis title to keep it clean
          panel.grid.major = element_line(color = "gray90", size = 0.5),  # Light grid lines for subtlety
          panel.grid.minor = element_blank(),  # Remove minor grid lines
          panel.border = element_blank(),  # Remove plot border for a more open look
          plot.margin = unit(c(1, 1, 1, 1), "cm"),  # Correct margin specification
          axis.line.x = element_line(color = "black", size = 1),  # Bold x-axis line
          axis.line.y = element_blank()  # Remove y-axis line for a cleaner look
        ) +
        scale_fill_gradient2(
          low = "#3B9AB2",    # Soft blue
          mid = "#E0E0E0",    # Light grey
          high = "#F21A00",   # Deep red
          midpoint = 0
        ) +
        scale_x_discrete(limits = rev(levels(coefficients_df$Feature))) +  # Ensure proper alignment of bars
        geom_hline(yintercept = 0, linetype = "solid", size = 1.2, color = "black")  # Bold horizontal line at y = 0
      
```

## Correlation Heatmap of Crude Oil Price, Exchange Rate, and Inflation Rate

```{r}
 # Calculate the correlation matrix for Crude Oil Price, Exchange Rate, and Inflation Rate
      cor_matrix <- cor(inf_rate[, c("Crude.Oil.Price", "Exchange Rate", "Inflation_Rate")])
   
      # Convert the correlation matrix into a long format
      cor_matrix_melted <- melt(cor_matrix)
```

## Visualise the Correlation

```{r}
# Plot the correlation heatmap
      ggplot(cor_matrix_melted, aes(Var1, Var2, fill = value)) +
        geom_tile() +
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0) +
        labs(
          title = "Correlation Heatmap of Crude Oil Price, Exchange Rate, and Inflation Rate",
          x = "Variables",
          y = "Variables",
          fill = "Correlation"
        ) +
        theme_minimal(base_size = 14) +
        theme(
          plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 14),
          panel.grid = element_blank()
        )
```

## Print the Correlation

```{r}
print(cor_matrix_melted)
```

## Correlation with Trend line

## Inflation Vs Crude oil Price

```{r}
# Plot Inflation Rate vs Crude Oil Price with a trend line
      ggplot(inf_rate, aes(x = Crude.Oil.Price, y = Inflation_Rate)) +
        geom_point(color = "blue") +                           # Plot the data points
        geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear trend line
        labs(title = "Inflation Rate vs Crude Oil Price",
             x = "Crude Oil Price", y = "Inflation Rate") +
        theme_minimal(base_size = 14)
```

## Inflation Vs Exchange Rate

```{r}

      # Plot Inflation Rate vs Exchange Rate with a trend line
      ggplot(inf_rate, aes(x = `Exchange Rate`, y = Inflation_Rate)) +
        geom_point(color = "green") +                           # Plot the data points
        geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear trend line
        labs(title = "Inflation Rate vs Exchange Rate",
             x = "Exchange Rate", y = "Inflation Rate") +
        theme_minimal(base_size = 14)
```

## Correlation Matrix for All Variables

```{r}
# Select all relevant columns including Inflation_Rate, CPI columns, Crude.Oil.Price, and Exchange Rate
      all_columns <- c("Inflation_Rate", grep("^CPI_", names(inf_rate), value = TRUE), "Crude.Oil.Price", "Exchange Rate")
      
      # Subset the data to include only the selected columns
      selected_data <- inf_rate[, all_columns]
      
      # Calculate the correlation matrix for all selected variables
      cor_matrix_all <- cor(selected_data, use = "complete.obs")
```

## Plot the Correlation Matrix

```{r}
 # Plot the correlation heatmap
      corrplot(cor_matrix_all, method = "color", type = "upper", 
               tl.cex = 0.8, tl.col = "black", addCoef.col = "white",
               title = "Correlation Matrix for All Variables", mar = c(0,0,1,0))
```

## Print the Matrix

```{r}
print(cor_matrix_all)
```

## Yearly average of inflation

```{r}
 # Extract the year from row names
      inf_rate$Year <- format(as.Date(rownames(inf_rate)), "%Y")
      
      # Replace 'InflationRate' with the actual column name from your dataset
      yearly_avg_inflation <- aggregate(Inflation_Rate ~ Year, data = inf_rate, FUN = mean)
```

## Visualise the Yearly Average

```{r}
ggplot(yearly_avg_inflation, aes(x = as.numeric(Year), y = Inflation_Rate)) +
        geom_line(color = "#228B22", size = 1.2) +
        geom_point(color = "red", size = 3) +
        geom_text(aes(label = round(Inflation_Rate, 2)), vjust = -0.8, size = 3.5) +
        labs(
          title = "Yearly Average Inflation Rate",
          x = "Year",
          y = "Average Inflation (%)"
        ) +
        scale_x_continuous(breaks = as.numeric(yearly_avg_inflation$Year)) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```

# Effect Analysis

```{r}
### Effect Analysis
      # If Date is row names (index), do this:
      inf_rate$Date <- as.Date(rownames(inf_rate))
      
      # Now perform slicing as usual
      before_covid <- inf_rate[inf_rate$Date < as.Date("2020-03-01"), "Inflation_Rate"]
      after_covid  <- inf_rate[inf_rate$Date >= as.Date("2020-03-01"), "Inflation_Rate"]
      
      cohen.d(before_covid, after_covid)
      t.test(before_covid, after_covid, var.equal = FALSE)
      
      sum(inf_rate$Date < as.Date("2020-03-01"))   # Count of before COVID
      sum(inf_rate$Date >= as.Date("2020-03-01"))  # Count of after COVID
```

## Visualise the

```{r}
 #BoxPlot
      boxplot(before_covid, after_covid, names = c("Before COVID", "After COVID"),
              col = c("lightblue", "lightgreen"),
              main = "Inflation Rate Before and After COVID")
```

## T-Test and Cohen Test

```{r}
t_test_result <- t.test(before_covid, after_covid, var.equal = FALSE)
      print(t_test_result)
      
      cohen_d_result <- cohen.d(before_covid, after_covid)
      print(cohen_d_result)
```

## Visualise the BoxPlot of Before and after COVID

```{r}

      
      # Create a new column for period
      inf_rate$Period <- ifelse(inf_rate$Date < as.Date("2020-03-01"), "Before COVID", "After COVID")
      
     
      # Boxplot
      ggplot(inf_rate, aes(x = Period, y = Inflation_Rate, fill = Period)) +
        geom_boxplot(width = 0.5, outlier.color = "red") +
        labs(title = "Inflation Rate in Nigeria Before and After COVID-19",
             x = "", y = "Inflation Rate (%)") +
        theme_minimal() +
        theme(legend.position = "none") +
        scale_fill_manual(values = c("Before COVID" = "#00BFC4", "After COVID" = "#F8766D"))
```

## Time Series Plot with Highlighted COVID Break Point

```{r}
##Time Series Plot with Highlighted COVID Break Point
      ggplot(inf_rate, aes(x = Date, y = Inflation_Rate)) +
        geom_line(color = "#2C3E50", size = 1) +  # Deep blue line
        geom_point(shape = 16, color = "#E67E22", size = 1.5) +  # Tiny orange circles
        geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "#E74C3C", size = 0.8) +
        annotate("text", x = as.Date("2020-04-01"), y = max(inf_rate$Inflation_Rate, na.rm = TRUE) + 1,
                 label = "COVID-19 Start", color = "#E74C3C", angle = 90, hjust = 0, size = 3.2, fontface = "italic") +
        labs(
          title = "📈 Monthly Inflation Rate in Nigeria (2008–2024)",
          subtitle = "Data Source: National Bureau of Statistics (NBS) and Central Bank of Nigeria (CBN)",
          x = "Year", y = "Inflation Rate (%)"
        ) +
        scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
        theme_minimal(base_size = 12) +
        theme(
          plot.title = element_text(face = "bold", size = 14, color = "#2C3E50"),
          plot.subtitle = element_text(size = 10, color = "gray40"),
          axis.title = element_text(face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.minor = element_blank()
        )
```

# CONVERT THE DATA SET TO TIME SERIES

```{r}
# Convert the log-transformed column to a time series object
      ts_online <- ts(inf_rate, start = c(2008, 1), frequency = 12)
```

## Inspect the data

```{r}
 print(head(ts_online))
      class(ts_online)
```

## CHECK FOR AND HANDLE OUTLIERS

```{r}
# Function to detect outliers using IQR
      detect_outliers <- function(ts_data) {
        Q1 <- quantile(ts_data, 0.25, na.rm = TRUE)
        Q3 <- quantile(ts_data, 0.75, na.rm = TRUE)
        IQR_value <- Q3 - Q1

        lower_bound <- Q1 - 1.5 * IQR_value
        upper_bound <- Q3 + 1.5 * IQR_value

        outliers <- which(ts_data < lower_bound | ts_data > upper_bound)
        return(outliers)
      }

      # Find outliers in target variable
      outliers_target <- detect_outliers(ts_online)

      # Print detected outliers
      print(outliers_target)
```

## Handle Outliers

```{r}
# Function to replace outliers with interpolated values
      handle_outliers <- function(ts_data, outliers) {
        ts_cleaned <- ts_data
        ts_cleaned[outliers] <- NA  # Replace outliers with NA
        ts_cleaned <- zoo::na.approx(ts_cleaned, rule = 2)  # Interpolate missing values
        return(ts_cleaned)
      }

      # Handle outliers in target variable
      ts_online_cleaned <- handle_outliers(ts_online, outliers_target)
```

## Plot of Crude Oil Price VS Exchange Rate Vs Inflation Rate

```{r}
 # Extract date from time index of ts object
      dates <- as.Date(as.yearmon(time(ts_online)))
      
      # Create a data frame with required columns
      df_combined <- tibble(
        Date = dates,
        Inflation_Rate = ts_online_cleaned[, "Inflation_Rate"],
        Crude.Oil.Price = ts_online_cleaned[, "Crude.Oil.Price"],
        `Exchange Rate` = ts_online_cleaned[, "Exchange Rate"]
      )
      
      # Reshape to long format for ggplot
      df_long <- df_combined %>%
        pivot_longer(cols = c(Inflation_Rate, Crude.Oil.Price, `Exchange Rate`),
                     names_to = "Variable", values_to = "Value")
      
      # Plot
      ggplot(df_long, aes(x = Date, y = Value, color = Variable)) +
        geom_line(size = 1) +
        labs(title = "Inflation Rate, Crude Oil Price & Exchange Rate (2008 - Present)",
             x = "Year", y = "Value", color = "Variable") +
        scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Reconstruct and Plot

```{r}
 ##Reconstructing the tS data
      ts_inf <- ts(ts_online, start = c(2008, 1), frequency = 12 )
      
      class(ts_inf) 
      print(head(ts_inf))
      
      ##Visualizing Before and After Outlier Handling
      
      autoplot(ts_inf)
```

## Decompose the Variables

```{r}
# Extract the specific time series as a vector from the matrix
      ts_inflation <- ts_inf[, "Inflation_Rate"]
      ts_crude <- ts_inf[, "Crude.Oil.Price"]
      ts_xchange <- ts_inf[, "Exchange Rate"]
      
      decomposed_inflation <- stl(ts(ts_inflation, frequency = 12), s.window = "periodic") # Seasonal decomposition
      decomposed_crude  <- stl(ts(ts_crude, frequency = 12), s.window = "periodic")
      decomposed_xchange <- stl(ts(ts_xchange, frequency = 12), s.window = "periodic")
```

## Plot the decomposition

```{r}
# Plot the decomposition
      autoplot(decomposed_inflation) +
        labs(title = "Inflation Rate",
             x = "Date") +
        theme_minimal()
      
      autoplot(decomposed_xchange) +
        labs(title = "Decomposition of Exchange Rate",
             x = "Date") +
        theme_minimal()
      
      autoplot(decomposed_crude) +
        labs(title = "Decomposition of Crude Oil Price",
             x = "Date") +
        theme_minimal()
      
```

## Run ADF Test to check for Stationarity

```{r}
# ADF test for Inflation Rate
      adf_inflation_rate <- adf.test(ts_inflation)
      
      # ADF test for Crude Oil Price
      adf_crude_oil <- adf.test(ts_crude)
      
      # ADF test for Exchange Rate
      adf_exchange_rate <- adf.test(ts_xchange)
```

## Print the result of the ADF Test

```{r}
# Print and interpret results
      cat("\nADF Test Result for Inflation Rate:\n")
      print(adf_inflation_rate)
      if (adf_inflation_rate$p.value < 0.05) {
        cat("=> Inflation Rate is stationary (p-value:", adf_inflation_rate$p.value, ")\n")
      } else {
        cat("=> Inflation Rate is NOT stationary (p-value:", adf_inflation_rate$p.value, ")\n")
      }
      
      cat("\nADF Test Result for Crude Oil Price:\n")
      print(adf_crude_oil)
      if (adf_crude_oil$p.value < 0.05) {
        cat("=> Crude Oil Price is stationary (p-value:", adf_crude_oil$p.value, ")\n")
      } else {
        cat("=> Crude Oil Price is NOT stationary (p-value:", adf_crude_oil$p.value, ")\n")
      }
      
      cat("\nADF Test Result for Exchange Rate:\n")
      print(adf_exchange_rate)
      if (adf_exchange_rate$p.value < 0.05) {
        cat("=> Exchange Rate is stationary (p-value:", adf_exchange_rate$p.value, ")\n")
      } else {
        cat("=> Exchange Rate is NOT stationary (p-value:", adf_exchange_rate$p.value, ")\n")
      }
      
```

## Applying NDIFFS to get the number of differencing needed

```{r}
 # Apply ndiffs to determine how many differences are needed for stationarity
      # Find the number of differences required to make the series stationary
      diffs_needed_inflation <- ndiffs(ts_inflation)
      diffs_needed_crude_oil <- ndiffs(ts_crude)
      diffs_needed_exchange_rate <- ndiffs(ts_xchange)
      
      # Print the number of differences needed for each time series
      print(diffs_needed_inflation)
      print(diffs_needed_crude_oil)
      print(diffs_needed_exchange_rate)
```

## Apply differencing based on the number of differences calculated

```{r}
 # Apply differencing based on the number of differences calculated
      differenced_inflation <- diff(ts_inflation, differences = diffs_needed_inflation)
      differenced_crude_oil <- diff(ts_crude, differences = diffs_needed_crude_oil)
      differenced_exchange_rate <- diff(ts_xchange, differences = diffs_needed_exchange_rate) 
      
      # Print the differenced time series
      print(differenced_inflation)
      print(differenced_crude_oil)
      print(differenced_exchange_rate)
```

## Check the length of the original and differenced time series

```{r}
# Check the length of the original and differenced time series
      length(ts_inf[, "Inflation_Rate"])
      length(differenced_inflation)
      
      length(ts_inf[, "Crude.Oil.Price"])
      length(differenced_crude_oil)
      
      length(ts_inf[, "Exchange Rate"])
      length(differenced_exchange_rate)
```

## Align inflation rate and crude oil with exchange rate

```{r}
differenced_inflation <- head(differenced_inflation, -1)
      differenced_crude_oil <- head(differenced_crude_oil, -1)
```

## Apply ADF test on the differenced time series

```{r}
# Apply ADF test on the differenced time series
      adf_differenced_inflation <- adf.test(differenced_inflation)
      adf_differenced_crude_oil <- adf.test(differenced_crude_oil)
      adf_differenced_exchange_rate <- adf.test(differenced_exchange_rate)
```

## Print the results

```{r}
# Print results for differenced time series
      cat("\nADF Test Result for Differenced Inflation Rate:\n")
      print(adf_differenced_inflation)
      
      cat("\nADF Test Result for Differenced Crude Oil Price:\n")
      print(adf_differenced_crude_oil)
      
      cat("\nADF Test Result for Differenced Exchange Rate:\n")
      print(adf_differenced_exchange_rate)
      
```

## Decompose the Differenced Time Series

```{r}

      # Step 1: Decompose the differenced time series using STL (Seasonal and Trend decomposition using LOESS)
      decomposed_differenced_inflation <- stl(differenced_inflation, s.window = "periodic", t.window = 12)
      decomposed_differenced_crude_oil <- stl(differenced_crude_oil, s.window = "periodic", t.window = 12)
      decomposed_differenced_exchange_rate <- stl(differenced_exchange_rate, s.window = "periodic", t.window = 12)
```

## Plot the Decomposition

```{r}
# Plot the decompositions using autoplot() for each differenced time series
      
      autoplot(decomposed_differenced_inflation) + 
        labs(title = "Decomposition of Differenced Inflation Rate", 
             subtitle = "Trend, Seasonal, and Remainder")
      
      autoplot(decomposed_differenced_crude_oil) + 
        labs(title = "Decomposition of Differenced Crude Oil Price", 
             subtitle = "Trend, Seasonal, and Remainder")
      
      autoplot(decomposed_differenced_exchange_rate) + 
        labs(title = "Decomposition of Differenced Exchange Rate", 
             subtitle = "Trend, Seasonal, and Remainder")
```

## Check lengths of differenced variables

```{r}
 # Check lengths of differenced variables
      length(differenced_inflation)  # Length of differenced inflation rate
      length(differenced_crude_oil)  # Length of differenced crude oil price
      length(differenced_exchange_rate)  # Length of differenced exchange rate
      
```

## Align Variables for Model fitting

```{r}
# 1. Align the target variable (differenced_inflation) after differencing
      inflation_aligned <- differenced_inflation  # Already length = 196
      class(inflation_aligned)
      # 2. Combine the differenced exogenous variable (Exchange Rate)
      exo_var <- differenced_exchange_rate  # Using only Exchange Rate as exogenous variable
      
      # Confirm lengths
      length(inflation_aligned)  # Should be 196
     length(exo_var)              # Should be 196 x 1 (since only Exchange Rate is used as exogenous variable)
      
     print(inflation_aligned)
```

## Split the dataset into Train and Test

```{r}
#SPLIT THE DATASET INTO TRAINING AND TESTING SETS
     # Split index (80% training, 20% testing)
     n <- length(inflation_aligned)  # 196
     split_index <- floor(0.8 * n)   # 156
     
     # Create training and testing sets
     inflation_train <- inflation_aligned[1:split_index]
     inflation_test  <- inflation_aligned[(split_index + 1):n]
     
     exo_train <- exo_var[1:split_index]
     exo_test  <- exo_var[(split_index + 1):n]
     
     # Convert target series to time series format (monthly frequency)
     inflation_train_ts <- ts(inflation_train, start = c(2008, 2), frequency = 12)
     inflation_test_ts  <- ts(inflation_test, start = c(2008 + floor((split_index + 1) / 12), ((split_index + 1) %% 12) + 1), frequency = 12)
     
     # Convert exogenous variable to time series format (monthly frequency)
     exo_train_ts <- ts(exo_train, start = c(2008, 2), frequency = 12)
     exo_test_ts  <- ts(exo_test, start = c(2008 + floor((split_index + 1) / 12), ((split_index + 1) %% 12) + 1), frequency = 12)
      
```

# FIT MODELS FOR FORECASTING

## SARIMAX MODE

```{r}
 # Use the whole time series (inflation_aligned) for training
     inflation_ts <- inflation_aligned
     
     # Assuming exo_var is also a time series object
     exo_ts <- exo_var
     
     # Fit a SARIMAX model (ARIMA with exogenous variables)
     sarimax_model <- auto.arima(inflation_ts, xreg = exo_ts, seasonal = TRUE)
     summary(sarimax_model)
     
     
          ##Model Diagnostics
     # Ljung-Box Test on residuals
     Ljung_Sarimax <- Box.test(residuals(sarimax_model), lag = 20, type = "Ljung-Box")
     
     # AIC and BIC
     Aic_Sarimax <- AIC(sarimax_model)
     BIC_Sarimax <- (sarimax_model)
     
     print(Ljung_Sarimax)
     print(Aic_Sarimax)
     print(BIC_Sarimax)
     
```

## Forecast with Exogenous Variable

```{r}
# Forecast the next 12 months (June 2024 to December 2025)
     forecast_horizon <- 12
     exo_future <- tail(exo_ts, 1)  # Get the last value of the exogenous variable
     exo_future <- rep(exo_future, forecast_horizon)  # Replicate for the forecast horizon 

# Forecast the inflation rate considering the exogenous variable
     sarimax_forecast_future <- forecast(sarimax_model, xreg = exo_future, h = forecast_horizon)
     # Last actual inflation value (May 2024) before the forecast (this should be the real value from the dataset)
     last_actual_value <- 33.69  # Adjust this as per the actual last value in your dataset
     
     # Get the differenced forecasted inflation values from SARIMAX forecast
     forecasted_differenced_values <- sarimax_forecast_future$mean
     
     # Reverse differencing: Add last actual value to the cumulative sum of differenced values
     inflation_forecasted_actuals <- last_actual_value + cumsum(forecasted_differenced_values)
     
     # Print the original inflation values after reversing differencing
     print(inflation_forecasted_actuals)
     
     # Assume inflation_forecasted_actuals are the forecasted inflation values and forecasted_dates are the corresponding dates
     forecasted_dates <- seq(as.Date("2024-05-01"), by = "month", length.out = length(inflation_forecasted_actuals))
     
     # Create a data frame with the forecasted dates and inflation values
     forecast_data <- data.frame(Date = forecasted_dates, Inflation = inflation_forecasted_actuals)
```

## Plot the Forecasted Values from May 2024 - April 2025

```{r}
# Plot using ggplot2 with values on top of the marks
     ggplot(forecast_data, aes(x = Date, y = Inflation)) +
       geom_line(color = "blue", size = 1) +  # Line plot
       geom_point(color = "red", size = 2) +  # Points on the line
       geom_text(aes(label = round(Inflation, 2)), vjust = -0.5, color = "black", size = 3) +  # Add text labels
       labs(title = "Inflation Rate Forecast (May 2024 - April 2025)",
            x = "Date", y = "Inflation Rate (%)") +
       theme_minimal() +
       theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
       scale_x_date(date_labels = "%b %Y", date_breaks = "1 month")  # Format the x-axis labels to display months and years
```

# SARIMA MODEL

## Fit the Model

```{r}
# Use the whole time series (inflation_aligned) for training
     inflation_ts <- inflation_aligned
     
     # Fit a SARIMA model 
     sarima_model <- auto.arima(inflation_ts, seasonal = TRUE)
     summary(sarima_model)
     
     
     ##Model Diagnostics
     # Ljung-Box Test on residuals
     Ljung_Sarima <- Box.test(residuals(sarima_model), lag = 20, type = "Ljung-Box")
     
     # AIC and BIC
     Aic_Sarima <- AIC(sarima_model)
     BIC_Sarima <- (sarima_model)
     
     print(Ljung_Sarima)
     print(Aic_Sarima)
     print(BIC_Sarima)
```

## Forecast with SARIMA

```{r}
 # Forecast the next 12 months (June 2024 to December 2025)
     forecast_horizon <- 12
    
      # Forecast the inflation rate considering the exogenous variable
     sarima_forecast_future <- forecast(sarima_model, h = forecast_horizon)
     # Last actual inflation value (May 2024) before the forecast (this should be the real value from the dataset)
     last_actual_value <- 33.69  # Adjust this as per the actual last value in your dataset
     
     # Get the differenced forecasted inflation values from SARIMAX forecast
     forecasted_differenced_values_sarima <- sarima_forecast_future$mean
     
     # Reverse differencing: Add last actual value to the cumulative sum of differenced values
     inflation_forecasted_actuals_sarima <- last_actual_value + cumsum(forecasted_differenced_values_sarima)
     
     # Print the original inflation values after reversing differencing
     print(inflation_forecasted_actuals_sarima)
     
     # Create a sequence of dates starting from June 2024 (or the start of your forecast) up to December 2025
     forecasted_dates <- seq(as.Date("2024-05-01"), by = "month", length.out = length(inflation_forecasted_actuals_sarima))
     
     # Create a data frame with forecasted dates and inflation values
     forecast_data <- data.frame(Date = forecasted_dates, Inflation = inflation_forecasted_actuals_sarima)
```

## Plot the Forecasted Values

```{r}
# Plot using ggplot2 with values on each point
     ggplot(forecast_data, aes(x = Date, y = Inflation)) +
       geom_line(color = "blue", size = 1) +  # Line plot
       geom_point(color = "red", size = 2) +  # Points on the line
       geom_text(aes(label = round(Inflation, 2)), vjust = -0.5, color = "black", size = 3) +  # Add values on each point
       labs(title = "Inflation Rate Forecast (May 2024 - April 2025)",
            x = "Date", y = "Inflation Rate (%)") +
       theme_minimal() +
       theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
       scale_x_date(date_labels = "%b %Y", date_breaks = "1 month")  # Format the x-axis labels to display months and years
```

## Granger Causality Test

```{r}
#Granger Causality Effect
# Granger Causality Test between Inflation Rate and Exchange Rate
     granger_inflation_exchange <- grangertest(Inflation_Rate ~ `Exchange Rate`, order = 2, data = ts_inf)
     
     # Granger Causality Test between Inflation Rate and Crude Oil Price
     granger_inflation_oil <- grangertest(Inflation_Rate ~ Crude.Oil.Price, order = 2, data = ts_inf)
     
     # Display the results of both tests
     granger_inflation_exchange
     granger_inflation_oil
```

# The End!
